---
title: "OtterDoc"
format: html
editor: visual
---

Downloading the data!

```{r}
#Libraries:
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lme4)
library(cowplot)
library(car)
library(ggeffects)
library(lubridate)

otterdat <- read_csv(here::here("data/mompupdata_all.xlsx - Sheet1.csv"))
momdat <- read_csv(here::here("data/mombday.csv"))

names(momdat)[names(momdat) == "Brthdate_est"] <- "mombday"
names(otterdat)[names(otterdat) == "Brthdate_est"] <- "pupbday"
names(momdat)[names(momdat) == "OTTERNO"] <- "Otterno"
```

Now that we have our datasets somewhat organized... I need to join them as well find accurate agecert values for the females..

Here, we will assign an estimated age using agecert while allowing some randomness

```{r}
#First change the birthdate year to a correct year form
momdat <- momdat %>%
  mutate(
    mombday = mdy(mombday)
  )

#Now bring the columns from momdat to otterdat
otterdat <- otterdat %>%
  left_join(momdat %>% 
              select(Otterno, mombday, AGECERT_mom),
            by = "Otterno")

#Now we can assign an age to mom's at time of pup's birth using
otterdat <- otterdat %>%
  mutate(
    mom_age = as.integer(interval(mombday, pupbday) / years(1))
  )


#Trying to get a better age estimate using agecert
otterdat <- otterdat %>%
  mutate(
    mom_age_estimated = case_when(
      AGECERT_mom == 4 ~ paste0(mom_age - 1, "-", mom_age + 1),
      AGECERT_mom == 5 ~ paste0(mom_age - 2, "-", mom_age + 2),
      AGECERT_mom == 6 ~ "adult (unknown age)",
      TRUE ~ NA_character_
    )
  )

#This just makes it all more confusing... 
```

So.. We an assigned age for mom's that really isn't helpful at all and we have our data all organized into otterdat now..

While it isn't really accurate.. we can try to use the estimated ages without age_cert to see what happens

Our first quick analysis on the correlation of how weaning success varies with mom's age

```{r}
otterdat %>%
  group_by(mom_age) %>%
  summarize(
    wean_success_rate = mean(Wean_success == TRUE, na.rm = TRUE),
    pup_survival_rate = mean(Mortality == FALSE, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(mom_age)

#Looks like there is definitely a clear case of older mom's have higher rates of pup survival and weaning:

ggplot(otterdat, aes(x = mom_age, fill = Wean_success)) +
  geom_bar(position = "fill") +
  labs(title = "Weaning Success by Mom's Age", y = "Proportion", x = "Mom's Age at Pup Birth") +
  scale_fill_manual(values = c("TRUE" = "green", "FALSE" = "red"))

```

So with that quick analysis.. maybe there is some correlation

Although I'm still not completely believing it.. so maybe what we do now is create a GLMM model for this

```{r}
#First I know I want random effect of year and individual so let's create a column for year
otterdat <- otterdat %>%
  mutate(year = year(pupbday))

#Then in order to measure wean sucess we want to turn Wean_success to an integer
otterdat <- otterdat %>%
  mutate(
    Wean_success = as.integer(Wean_success == TRUE)
  )

#Now we can build the model using a logistic regression mixed model:

otter_mod <- glmer(Wean_success ~ mom_age + (1 | Otterno) + (1 | year),
                     data = otterdat,
                     family = binomial,
                     control = glmerControl(optimizer = "bobyqa"))

summary(otter_mod)
```

Right away... look's like there's no significance however detecting significance between individuals ages is hard to do.. so let's try to add some new variables to express our findings!

First we can group Mom Age into age classes and turn pup sex into a factor

```{r}
otterdat2 <- otterdat %>%
  mutate(
    # Mom age group
    mom_age_group = case_when(
      mom_age < 4 ~ "young",
      mom_age >= 4 & mom_age <= 8 ~ "prime",
      mom_age > 8 ~ "old",
      TRUE ~ NA_character_
    ),
    mom_age_group = factor(mom_age_group, levels = c("young", "prime", "old")),

    # Pup sex (factor)
    SEX = factor(SEX, levels = c("m", "f", "u")),

    # Binary outcome for weaning success
    wean_success = as.integer(Wean_success == TRUE),

    # Binary outcome for pup sex (for model 2)
    sex_bin = ifelse(SEX == "m", 1, 0)
  ) %>%
   filter(!is.na(mom_age_group), !is.na(SEX), !is.na(wean_success), !is.na(mom_age), !is.na(pupbday))
```

Now that our data is all organized we can create our models

For our first question: Does mom's age group/class affect weaning success?

```{r}
age_mod <- glmer(
  wean_success ~ mom_age_group + (1 | Otterno) + (1 | year),
  data = otterdat2,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa")
)

summary(age_mod)

#After, create a customprediction grid:
wean_grid <- expand_grid(
  mom_age_group = factor(c("young", "prime", "old"), levels = c("young", "prime", "old")))
  
#Then create a prediction for population levels
wean_pred <- wean_grid %>%
  mutate(predicted = predict(
    age_mod,
    newdata = wean_grid,
    re.form = NA,
    type = "response"
  ))

#Add CI
wean_pred_ci <- ggpredict(age_mod, terms = "mom_age_group") %>% as.data.frame()

#Then create a summary for the data:
wean_summ <- otterdat2 %>%
  group_by(mom_age_group) %>%
  summarise(
    mean_success = mean(wean_success, na.rm = TRUE),
    se_success = sd(wean_success, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

#Add some jitter:
wean_summ <- wean_summ %>%
  mutate(x_jitter = as.numeric(mom_age_group) + 0.1)

#fix a problem:
wean_pred_ci$x <- factor(c("young", "prime", "old"), levels = levels(wean_summ$mom_age_group))

#Now we can hopefully graph this! 

#Plot
ggplot() +
  geom_pointrange(
    data = wean_summ,
    aes(
      x = mom_age_group,
      y = mean_success,
      ymin = mean_success - 1.96 * se_success,
      ymax = mean_success + 1.96 * se_success
    ),
    position = position_dodge(width = 0.5),
    shape = 21,
    fill = "#D5A6F6",
    color = "#8E44AD",
    size = 1.2,
    show.legend = FALSE
  ) +
  geom_line(
    data = wean_pred_ci,
    aes(x = x, y = predicted, group = 1),
    color = "#6DD5B3",
    linewidth = 1.5
  ) +
  geom_ribbon(
    data = wean_pred_ci,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2,
    fill = "#6DD5B3"
  ) +
  scale_x_discrete(name = "Mom Age Group") +
  theme_minimal() +
  labs(
    title = "Weaning Success by Mom Age Class",
    y = "Probability of Weaning Success"
  ) +
  theme(
    axis.title = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "italic", hjust = 0.5, size = 16)
  )


```

Now for Question 2:

How Does Mom's age group impact \# of male pups produced:

```{r}
#I have to join young and prime together since all the young pups are female
otterdat2$mom_age_group2 <- fct_collapse(otterdat2$mom_age_group,
                                         young = c("young", "prime"),
                                         old = "old")
#Create new model:
sex_mod <- glmer(
  sex_bin ~ mom_age_group2 + (1 | Otterno),
  data = otterdat2,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa")
)

summary(sex_mod)

#Again no significance
sex_pred_ci <- ggpredict(sex_mod, terms = "mom_age_group2") %>% as.data.frame()

sex_summ <- otterdat2 %>%
  group_by(mom_age_group2) %>%
  summarise(
    mean_sex = mean(as.numeric(as.character(sex_bin)), na.rm = TRUE),
    se_sex = sd(as.numeric(as.character(sex_bin)), na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

sex_summ <- sex_summ %>%
  mutate(x_jitter = as.numeric(mom_age_group2) + 0.1)
#And then plot:
ggplot() +
  # Observed means with 95% CI bars
  geom_pointrange(
    data = sex_summ,
    aes(
      x = mom_age_group2,
      y = mean_sex,
      ymin = mean_sex - 1.96 * se_sex,
      ymax = mean_sex + 1.96 * se_sex
    ),
    shape = 21, fill = "#D5A6F6", color = "#8E44AD", size = 1.2
  ) +
  # Predicted probabilities and confidence ribbons
  geom_line(
    data = sex_pred_ci,
    aes(x = x, y = predicted, group = 1),
    color = "#6DD5B3",
    size = 1.5
  ) +
  geom_ribbon(
    data = sex_pred_ci,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2,
    fill = "#6DD5B3"
  ) +
  scale_x_discrete(name = "Mom Age Group") +
  theme_minimal() +
  labs(
    title = "Probability of Producing Male Pups",
    subtitle = "Model Predictions with Observed Means",
    y = "Probability of Male Pup"
  ) +
  theme(
    axis.title = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "italic", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "#555555")
  )
```

Now our final question:

How do both of these factors interacting with each other impact weaning success

```{r}
otterdat2 <- otterdat2 %>%
  mutate(
    mom_age_group2 = factor(mom_age_group2, levels = c("young", "old")),
    sex_bin = factor(sex_bin, levels = c(0, 1))  # 0 = female, 1 = male, or vice versa as you set
  )
#Make sex_bin a factor:
otterdat2 <- otterdat2 %>%
  filter(!is.na(sex_bin)) %>%        # remove NAs
  filter(sex_bin %in% c(0, 1)) %>%   # keep only valid levels
  mutate(sex_bin = factor(sex_bin, levels = c(0, 1)))

#Create model for interaction:
int_mod <- glmer(
  wean_success ~ mom_age_group2 * sex_bin + (1 | Otterno) + (1 | year),
  data = otterdat2,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa")
)
summary(int_mod)

int_pred_ci <- ggpredict(int_mod, terms = c("mom_age_group2", "sex_bin")) %>% as.data.frame()


int_summ <- otterdat2 %>%
  group_by(mom_age_group2, sex_bin) %>%
  summarise(
    mean_success = mean(wean_success, na.rm = TRUE),
    se_success = sd(wean_success, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )


#Final plot!
ggplot() +
  # Observed points with error bars
  geom_pointrange(
    data = int_summ,
    aes(
      x = mom_age_group2,
      y = mean_success,
      ymin = mean_success - 1.96 * se_success,
      ymax = mean_success + 1.96 * se_success,
      fill = factor(sex_bin),
      color = factor(sex_bin)
    ),
    position = position_dodge(width = 0.5),
    shape = 21,
    size = 1.2
  ) +
  # Predicted lines
  geom_line(
    data = int_pred_ci,
    aes(x = x, y = predicted, group = group, color = group),
    position = position_dodge(width = 0.5),
    size = 1.5
  ) +
  # Confidence ribbons
  geom_ribbon(
    data = int_pred_ci,
    aes(x = x, ymin = conf.low, ymax = conf.high, group = group, fill = group),
    alpha = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  scale_x_discrete(name = "Mom Age Group") +
  scale_color_manual(name = "Pup Sex", values = c("#D5A6F6", "#1B9E77"), labels = c("Female", "Male")) +
  scale_fill_manual(name = "Pup Sex", values = c("#D5A6F6", "#1B9E77"), labels = c("Female", "Male")) +
  theme_minimal() +
  labs(
    title = "Weaning Success across Age and Sex",
    y = "Probability of Weaning Success"
  ) +
  theme(
    axis.title = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "italic", hjust = 0.5, size = 16)
    )
  
```
